---
title: 'Data Wrangling: Moorea Coral Reef LTER Fish Data'
author: "Julianna Renzi"
date: "2025-06-18"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load packages}
# Clear environment for maximum reproducibility
rm(list=ls())

# load librarian (package for installing/loading packages)
if (!require("librarian")) install.packages("librarian")

# Load other necessary libraries
librarian::shelf(here, # relative file paths
                 tidyverse # data wrangling
)
```

# Bring in data

Data downloaded from: https://portal.edirepository.org/nis/mapbrowse?scope=knb-lter-mcr&identifier=6&revision=64 on June 18, 2025 from "Annual Fish Surveys" (published on April 11, 2025)

```{r load data}
# raw fish counts
mcr_fish_1 <- read_csv(here("CSVs", "data_raw", "MCR_LTER_Annual_Fish_Survey_20250324.csv")) 

# taxa table where I cleaned up fish taxonomy and taxonomic resolution
mcr_taxa <- read_csv(here("CSVs", "taxa_tables", "mcr_fish_taxa_annotated.csv"))
```

**Some considerations wrt these data:**
- Biomass NA value is -1
- Biomass is reported in grams
- Will need to merge the two swaths (5 m and 1 m) to get total biomass estimates (both x 50 m)
- Will need to make decisions about spp not identified to the spp level
- "no fish observed" is an option
- Might want to remove sharks and very large transient fish, given Segrestin et al. 2024

# Account for zeros

How many instances are there of no fish observed?

```{r}
mcr_fish_1 %>% 
  filter(Taxonomy == "No fish observed") %>% dim() # two
```

Make "no fish observed" zeros with actual fish name (assign it to the first fish alphabetically)

```{r}
mcr_fish_1 %>% 
  mutate(Count = case_when(Taxonomy == "No fish observed" ~ 0,
                           TRUE ~ Count),
         Biomass = case_when(Taxonomy == "No fish observed" ~ 0,
                             TRUE ~ Biomass),
         Taxonomy = case_when(Taxonomy == "No fish observed" ~ "Abudefduf septemfasciatus",
                              TRUE ~ Taxonomy)) -> mcr_fish_2
```

# Biomass cleaning

We'll need biomass for getting fish community stability, which means we can only use fishes with biomass estimates.

How many observations don't have biomass estimates?

```{r}
mcr_fish_2 %>% 
  filter(Biomass == -1) %>% 
  summarize(n_individuals = sum(Count),
            n_rows = n()) 
```

See that there are 1,092 fish on 96 rows that don't have biomass estimates. The vast majority of these don't have length estimates (-1 for total length), although is "Teleostei unidentified" 

```{r}
mcr_fish_2 %>% 
  summarize(n_individuals = sum(Count),
            n_rows = n()) 
```

See that those 1,092 are just 0.24% of the 455,158 total individuals observed--so we will go ahead and filter these out.

```{r filter nonpositive biomass}
mcr_fish_2 %>% 
  filter(Biomass != -1) -> mcr_fish_3
```

# Taxa cleaning

Also want to clean up the taxonomy so that each taxon is unique. Use the table where I assigned taxonomic designations manually. First make sure we have all the taxa accounted for:

```{r}
mcr_fish_3 %>%
  left_join(mcr_taxa) %>% 
  filter(is.na(Taxonomy_updated)) %>% 
  select(Taxonomy) %>% 
  unique() # all accounted for (no NA's)
```

See how many we lose if we filter out non-specific taxonomy:

```{r}
mcr_fish_3 %>%
  left_join(mcr_taxa) %>% 
  group_by(id_confidence) %>% 
  summarize(abundance = sum(Count)) 
```

See that we lose less than 0.05% of observations if we filter out the vague IDs. Let's go ahead and do that (see Andy's notes below). We'll also update the taxonomy to deal with cf species/etc.

Andy Brooks notes that: 
- There is a unique description for every Blennie spp 6, etc. (is unique)
- Unknown Scaridae does not have a unique description
- They're such small numbers, probably fine to just exclude them; the most conservative thing is to just chuck them > also no way to do biomass
- Also FYI biomass is way more stable than abundance
- Andy also thinks it's fine to count the "cf" species as the species they are recorded as; sometimes these are for spp that are not even described (but reeeeally similar - 
length/weight probably the same; trophic is also probably the same); these differences are more regional 

```{r filter vague taxonomy}
mcr_fish_3 %>%
  left_join(mcr_taxa) %>% 
  # only keep well identified taxa
  filter(id_confidence == 1) %>% 
  # update taxonomy if needed
  select(-Taxonomy) %>% 
  rename(Taxonomy = Taxonomy_updated) -> mcr_fish_4
```

**Notes on more wide-ranging taxa**:

If we want to filter out sharks, could filter subdivision != "Selachii"

Other potentially wide ranging taxa:
- Caranx genus
- Gnathanodon genus (trevally)
- Carangoides ferdau (trevally)
- Carangoides orthogrammus (trevally)
- Tuna (subfamily Scombrinae, Tribe Thunnini)
- Barracuda (Sphyraena qenie, Sphyraena barracuda)
- Remoras? (e.g., Echeneis naucrates)
- Rays? (e.g., Pateobatis fai, Aetobatus ocellatus)
- Hound needlefish? (Tylosurus crocodilus)
- Jobfish? (Aprion virescens)
- Milkfish? (Chanos chanos)
- Batfish? (Platax orbicularis)
- Queenfish? (Scomberoides lysan)
- Flyingfish? (Cypselurus simus)
- Others! Stopped here (in order of most abundant fish)

# Make taxonomy list to save separately

```{r}
mcr_fish_4 %>% 
  group_by(Taxonomy, taxon_resolution, Coarse_Trophic, Fine_Trophic) %>% 
  summarize(N_rows = n()) %>% 
  # get rid of bad duplicate
  filter(!(Taxonomy == "Abudefduf septemfasciatus" & Coarse_Trophic == "na")) %>% 
  write_csv(here("CSVs", "taxa_tables", "mcr_fish_trophic_designations.csv"))
  
```


# QA/QC

Make sure all transects/swaths accounted for each year:

```{r}
mcr_fish_4 %>% 
  group_by(Year, Site, Habitat, Transect, Swath) %>% 
  summarize(n = n()) %>% 
  group_by(Year, Site) %>% 
  summarize(n = n()) %>% view()
```

Look at temporal coverage:

```{r}
mcr_fish_4 %>% 
  group_by(Year, Site, Habitat, Transect, Swath) %>% 
  summarize(n = n()) %>% 
  group_by(Year, Site) %>% 
  summarize(n = n()) %>% 
  group_by(Year) %>% 
  summarize(n_site = n()) %>% view()
```

See that all 6 sites were done in completion each year with no gaps

# Adjust for swaths

```{r get in terms of the 5 x 50 m transect}
mcr_fish_4 %>% 
  # get biomass for 5 m swath (can divide by 250 if want per m^2)
  mutate(biomass_g_per_5x50_m2 = case_when(Swath == 1 ~ Biomass*5, 
                                           Swath == 5 ~ Biomass,
                                           TRUE ~ -9999)) -> mcr_fish_5
```

# Format

Make a nicer format for the working group

```{r}
mcr_fish_5 %>% 
  mutate(
    dataset = "mcr",
    site = str_to_lower(Site),
    transect = Transect,
    swath = Swath,
    habitat = str_to_lower(Habitat),
    year = Year,
    month = month(as.Date(Date)),
    day = day(as.Date(Date)),
    plot = paste0(site, "_", habitat), 
    subplot = paste0(Transect, "_", Swath, "m"),
    biomass_g_per_5x50_m2 = biomass_g_per_5x50_m2,
    taxon_name = Taxonomy
  ) %>% 
  select(dataset, site, transect, swath, habitat, year, month, day, plot, subplot, taxon_name, biomass_g_per_5x50_m2) -> mcr_fish_6
```

Check for NA's

```{r}
unique(is.na(mcr_fish_6))
```

Looks good.

# Write CSV

```{r}
write_csv(mcr_fish_6, here("CSVs", "data_cleaned", "mcr_fish_biomass_5x50m2.csv"))
```

