---
title: 'Data Wrangling: Santa Barbara Coastal LTER Fish Data'
author: "Julianna Renzi"
date: "2025-06-18"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load packages}
# Clear environment for maximum reproducibility
rm(list=ls())

# load librarian (package for installing/loading packages)
if (!require("librarian")) install.packages("librarian")

# Load other necessary libraries
librarian::shelf(here, # relative file paths
                 tidyverse # data wrangling
)
```

# Bring in data

Data downloaded from: https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-sbc.50.17 on June 18, 2025. The dataset was published on 9/13/2024.

```{r load data}
# raw data from EDI
sbc_fish_1 <- read_csv(here("data_raw", "Annual_All_Species_Biomass_at_transect_20240823.csv"))

# taxa table where I cleaned up fish taxonomy and taxonomic resolution
sbc_taxa <- read_csv(here("taxa_tables", "sbc_taxa_annotated.csv"))
```

**Some considerations wrt these data:**
- These data are from permanent transects at 11 sites
- BULL, CARP, NAPL started in 2000, others in 2001
- Two Santa Cruz sites began in summer 2004
- 2-8 transects/site; some transects added later
- Transect = 40 m x 2 m fixed plot
- Transects 3, 5, 6, 7, 8 at IVEE were added in 2011
- Transects are surveyed annually in the summer
- -99999 means the value was not recorded or not available 
- Reef fish are those 2 m from the benthos
- IMPORTANT: the accuracy of sampling fish may vary with water clarity and *data collected during sampling events when horizontal visibility was < 2 m should be used with caution* (although some groups leave these in--we can decide what meets our needs best)

# Account for zeros

Want to just get fish, but want to make sure we have zeros for where no fish were observed. Get a full transect/site/year list:

```{r}
sbc_fish_1 %>% 
  group_by(YEAR, MONTH, DATE, SITE, TRANSECT) %>% 
  summarize(n = n()) %>% 
  select(-n) -> transect_list
```

See if there are any transects without fish:

```{r}
sbc_fish_1 %>% 
  filter(COARSE_GROUPING == "FISH") %>% 
  group_by(YEAR, MONTH, DATE, SITE, TRANSECT) %>% 
  summarize(n = n()) -> fish_transect_list
```

Are they the same length?

```{r}
dim(transect_list)[1] == dim(fish_transect_list)[1]
```

Yes! All transects had at least 1 fish on them

# Biomass cleaning

We'll need biomass for getting fish community stability, which means we can only use fishes with biomass estimates. How many observations don't have biomass estimates?

Note:
- **WM_GM2** = Areal wet mass of individuals or colony, including chitonous or calcareous structures (g/m2)
- **DRY_GM2** = Dry mass density derived from estimates of density or percent cover and laboratory estimates of taxa dry mass (g/m^2)

```{r}
sbc_fish_1 %>% 
  filter(COARSE_GROUPING == "FISH") %>% 
  filter(DRY_GM2 < 0) %>% dim() # 580 rows have NAs for dry mass

sbc_fish_1 %>% 
  filter(COARSE_GROUPING == "FISH") %>% 
  filter(WM_GM2 < 0) %>% dim() # 580 rows have NAs for wet mass
```

See that all NA's are from 2004 and 2005. Perhaps then we start in 2006?

```{r filter out years before 2006}
sbc_fish_1 %>% 
  filter(COARSE_GROUPING == "FISH") %>% 
  filter(YEAR > 2005) -> sbc_fish_2
```

# Taxonomy cleaning

Can see that there are no unidentifiable fish taxa:

```{r}
sbc_fish_2 %>% 
  left_join(sbc_taxa) %>% 
  filter(id_confidence == 0) %>% dim()
```

# Temporal coverage

```{r}
sbc_fish_2 %>% 
  group_by(YEAR, SITE, TRANSECT) %>% 
  # arbitrary summary thing here to get 1 value per entry
  summarize(total = sum(WM_GM2)) %>% 
  group_by(YEAR, SITE) %>% 
  summarize(n_transects = n()) %>% 
  pivot_wider(names_from = SITE,
              values_from = n_transects
              ) -> transect_reps
```

Can see IVEE is the only one that has irregular transect numbers

# Format

Make a nicer format for the working group

```{r}
sbc_fish_2 %>% 
  mutate(
    dataset = "sbc",
    site = str_to_lower(SITE),
    transect = TRANSECT,
    year = YEAR,
    month = MONTH,
    plot = site,
    day = day(as.Date(DATE)),
    biomass_wet_g_per_m2 = WM_GM2,
    biomass_dry_g_per_m2 = DRY_GM2,
    taxon_name = SCIENTIFIC_NAME, 
    common_name = COMMON_NAME
  ) %>% 
  select(dataset, site, transect, year, month, day, plot, taxon_name, common_name, biomass_wet_g_per_m2, biomass_dry_g_per_m2) -> sbc_fish_3
```

Check for NA's

```{r}
unique(is.na(sbc_fish_3))
```

Looks good.

# Write CSV

```{r}
write_csv(sbc_fish_3, here("data_cleaned", "sbc_fish_biomass.csv"))
```
